How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar

### 2020-05-24 docs/manuscript.Rmd
fixing scripts & adding raincloud esque plots

```{r setupmain, include=FALSE}
## Libraries
library(knitr)
library(tidyverse)
library(data.table)
library(rgdal)
library(glue)
library(here)
library(magrittr)

@@

Using the approach described in @weissGlobalMapTravel2018, we used GPS points of ARMC to estimate the travel time to the nearest ARMC (i.e. a clinic provisioning PEP) for the country of Madagascar at an approximately 1 x 1 km scale. We then extracted the mean of travel times at the district and commune (the administrative unit below the district) level, as well as the mean of travel times weighted by the population in that grid cell. We also compared these metrics to euclidean distance, but taking the minimum distance from the administrative unit centroid to an ARMC. We used administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs.

@@

We validated this approach to estimating form undersubmission by comparing estimated vial demand given the total reported bites (corrected for under-submission) for the years 2014-2017 to actual vials provisioned to clinics over this same period. Vial demand was simulated under simplified assumptions of PEP administration and adherence[@rajeevHealthcareUtilizationProvisioning2018], assuming that patients reported randomly across the year and then completed doses 3, 7, and 28 days subsequent from the date reported. For this analysis, we look at the range of vial estimates assuming that all patients complete 3 vs. 4 doses total. During this period, the Thai Red Cross Intradermal regimen was used across Madagascar, with 0.2 mL administered per patient at each visit. Vials can be shared within a day between two patients, resulting in 0.1 background wastage per vial shared, plus any additional wasteage from unused doses discarded at the end of the day. We did not have the date or month when vials were provisioned, so we compared the totals across this four year period.

@@

## Modeling reported bites as a function of travel time to closest ARMC {#methods-bitemods}
We modeled the number of reported bites as a function of travel time ($T$) in a poisson regression framework as follows: 
$$ \mu_i = e^{(\beta_{t}T_i \ + \ \beta_0 \ )}P_i$$
$$ y_i = Poisson(\mu_i) $$
where $y_i$ is our data on average number of bites reported to a clinic annually and $\mu_i$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC ($T_i$) and the human population size ($P_i$) (an offset which scales the incidence to expected number of bites) for a given source location (district or commune). We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level. In order to more directly compare estimates between these two datasets, we also modeled the national data with a commune-level travel time covariate ($T_{j}$) so that:

@@

We tested whether the catchment random effect was capturing overdispersion rather than any catchment specific effects by extending these models to incorporate an overdispersion parameter, $\sigma_e$ [@gelman2006data]:

@@

We fit all models in a Bayesian regression framework via MCMC using the R package ‘rjags’. We used a weakly informative prior for the intercept of the models centered around the mean of the bite incidence for the given data set, and for all other parameters we centered the priors around zero ($norm(0, 10-6)$). For the variance terms ($\sigma_0$ and $\sigma_e$), we set uniform priors ($unif(0, 10)$). We calculated the Deviance Information Criterion (DIC, a metric of model fit to data) for each candidate model as well as potential scale reduction factors (psrf, a metric of model convergence) for each covariate in the model and the multivariate psrf for the whole model [@gelman2006data].

To test how well our models predicted the data, we sampled parameter estimates from the posterior distribution for each model to generate predictions to the data. In addition, we used the models to predict the data that were not used to fit the model (i.e. we used the estimates from models fitted to the national data to predict the Moramanga data, and used estimates from models fitted to the Moramanga data to predict the national data). Finally, to check how correcting for incomplete submission of forms affected our modeling results, we fitted our final models to the raw data uncorrected for submission (i.e. assuming that data was completely reported) and with a lower cut-off for identifying periods of undersubmission (7 days vs. 15 days, resulting in higher estimates of undersubmission and accordingly bite incidence).

@@

For $E_i$, we center the distribution at the lower end of our estimate of exposure incidence from the Moramanga District (42 exposures/100,000 persons). The range is estimated using the approach used in [@whorabiesmodelingconsortiumPotentialImpactImproved2019] where rabies incidence in the dog population is estimated to be approximately 1% annually in the absence of dog vaccination, and that on average a rabid dog exposes 0.39 persons. These assumptions are derived from modeling and long term contact tracing data from the Serengeti District, Tanzania [@hampsonRabiesExposuresPostExposure2008; @hampsonSurveillanceEstablishElimination2016; @whorabiesmodelingconsortiumPotentialImpactImproved2019]. We then take the range of exposure incidence as the probability of exposure multiplied by the number of expected rabies cases in dogs assuming 1% incidence and the estimate of the dog population using a range of human to dog ratios (HDRs) of 5 (upper estimate) and 25 (lower estimate). As there is little data on dog populations in Madagascar [@andriamandimbySurveillanceControlRabies2013], this range of HDRs encompasses a wide range of observed HDRs across Africa [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating], as well as recent subnational estimates from Madagascar [@leblancclaireRabiesMadagascarThreepronged2019], and also generates similar human exposure incidences as observed previously across Africa [CITE!!]. See figure S4 for further details.

@@

$B_i$, the number of total reported bites, is the mean predicted number of bites from the models of bite incidence as a function of travel times. To get $R_i$, the number of reported rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures, drawn from a triangular distribution with range between 0.2 - 0.6. So that rabid reported bites cannot exceed the total expected number of human rabies exposures or a maximum reporting (as we know that even with minimal travel times, people may not report for PEP for other reasons), we constrain $p_{rabid}$ as follows:
$$ p_{rabid}=\begin{cases} x, & \text{if}\ \frac{E_i\rho_{max}}{B_i} > x \\ \frac{E_i\rho_{max}}{B_i}, &\text{otherwise} \end{cases}$$
where $\rho_{max}$ is the maximum reporting. We estimate $\rho_{max}$ from the Moramanga ARMC data for the commune of Moramanga Ville, the closest commune to the ARMC (average of 3.12 minutes travel time to the clinic), and where contact tracing data estimated that approximately 2% of rabies exposure go unreported [@rajeevHealthcareUtilizationProvisioning2018]. We assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies. 
We subtract $R_i$ from $E_i$, the number of total human rabies exposures drawn from a triangular distribution with range of 15.6 - 76 exposures per 100,000 persons, to get the number of unreported exposures ($U_i$). Deaths due to rabies in each administrative unit ($D_i$) are then estimated by mutliplying $D_i$ times $p_{death}$.
Deaths averted are calculated similarly as the reported rabies exposures $R_i$ that would have died in the absence of PEP given $p_{death}$ (i.e $A_ip_{death}$). Table 1 lists all parameter values and sources used in the decision tree. \

@@

```{r ncsbs}
## candidate points
csbs <- read.csv(here("data/processed/clinics/csb2.csv"), stringsAsFactors = FALSE)
clin_per_comm <- read.csv(here("data/processed/clinics/clinic_per_comm.csv"), 
                          stringsAsFactors = FALSE)
```

We developed a framework to rank clinics by how much they could improve access for people currently underserved (> 3 hours away from any clinic in the country) and to estimate the incremental reduction in burden as PEP is expanded to these clinics. We use the approximate locations of all existing public CSB II (centre de sante niveau II, the level below district hospitals, n = `r nrow(csbs)`) as locations of potential ARMC.

Starting with the current locations, we added one clinic at a time, calculating the proportion of people living < 3 hours away from any clinic for the country. We added the clinic which minimized this metric and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially. In this iterative process, we also exclude any clinics that once added only reduced travel times below the 3 hour threshold for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` persons). Finally, we estimate the incremental reduction in burden as each clinic is added using the burden framework described in the previous section.

We also explore three additional scenarios: 
1) Each district has one clinic (the CSB II in the highest population density grid cell compared to other potential clinics) provision PEP (N = 114 clinics). 
2) Each commune has one clinic (the CSB II, or when there are no CSB II in a commune the health post or dispensary, in the highest density grid cell compared to other potential clinics) provision PEP (N = `r nrow(clinic_per_comm)`. Note that not all communes (N = `r length(mada_communes)`) have a clinic).
3) All CSB II in the country provision PEP (N = `r nrow(csbs)` clinics).

We estimate how expansion to additional clinics may impact vaccine demand. Specifically, we use our model-predicted estimates of annual bites at the district and commune level and then aggregate these to the clinic level. As clinics are added, multiple clinics may serve a single district or commune (S5.4). Therefore, we allocated bites to each clinic according to the proportion of the population in each administrative unit for which it was the closest. For each clinic, we simulate throughput by randomly assigning patient presentation dates, and then assume perfect compliance (i.e. all patients reporting 3 and 7 days after their first dose) to generate dates of subsequent doses. We use these dates to estimate vial usage given routine vial sharing practices in Madagascar as described previously, but under the assumption of adoption of the new abridged intradermal regimen recommended by WHO (2 injection sites on day 0, 3, and 7, [@tarantolaIntradermalRabiesPostexposure2018]). We take the mean of 1000 simulations of throughput for each catchment and set of clinics.

## Sensitivity analysis {#methods-sensitivity}
To test the effect of our model assumptions on estimates of rabies burden and vial demand, we did a univariate sensitivity analyses of both parameters from the models of bite incidence and the decision tree parameters. For each parameter in the decision tree, we simulated predcitions fixing estimates to the upper and lower range of each parameter, while keeping the other parameters unchanged (Table S6.1). We did the same for parameter estimates from the models of reported bite incidence, but rather fixed these estimates at the upper and lower 95% quantiles of their posterior distributions (Table S6.2). We also look at how systematic variation in rabies incidence with human population size could affect our estimates of burden. Specifically, if we believe that human to dog ratios are positively correlated with human populations (i.e. if dog ownership/populations are higher in more populated, urban areas), then we might expect higher rabies exposure incidence within our range as population size increases. Alternatively, if we believe that the human to dog ratio is inversely correlated with human population size (i.e. if dog ownership is more common in less populated, rural areas), we would expect rabies exposure incidence within our range to scale negatively with human population size. We use scaling factors which scale incidence either positively or negatively with the population sizes observed at the district and commune levels, but are still constrained within the range of exposure incidence used in the main analyses (15.6 - 76 exposures per 100,000 persons, see Fig S6.4).

## Data and code 
All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the [supplementary references]{#srefs}. All processed data, code, and outputs are archived at this zenodo repository (to be archived upon submission) and maintained at this github repository. The raw bite patient data at the national level are maintained in a database by IPM. This database was last queried on September 19, 2019 for these analyses. Anonymized raw data from this database are available from IPM following insitutional data transfer protocols. Anonymized raw data collected from the Moramanga District were retrieved from the Wise Monkey Portal (http://www.wisemonkeyfoundation.org/) on the same date and are shared in the archived repository.

@@

Vial demand also increases (Fig 7B), and more vials are needed per averted death as PEP access is expanded (Fig 7C). Our model predictes an increase from `r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 0]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 0]), 2)` vials under current clinic provisioning and with the abridged Thai Red cross regimen (i.e. visits on days 0, 3, 7), to ~`r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 250]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 250]), 2)` vials with 250 clinics, and approximately `r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` vials if all CSB IIs were to provision PEP. Importantly, clinic throughput decreases, as well, with clinics seeing fewer patients per day (S5.7), as reported bite incidence saturates (Fig S5.5), and instead there are shifts in which clinics bites are reported to.

We also look at two additional scenarios: one CSB II provisions PEP for each district (ARMC per district, N = 114 clinics) and one where one clinic (either a CSB II, CSB I, or dispensary) provisions PEP to each commune (ARMC per commune, N = `r nrow(clin_per_comm)` as all communes have one clinic). These expansion scenarios result in similar estimates of impacts on reducing human rabies deaths and increasing vial demand compared to adding the same number of clinics through our stepwise approach.

@@

## Main findings
Travel time to the closest ARMC was a strong predictor of reported bite incidence in Madagascar. Using an adapted decision tree framework, we find that the burden of rabies is likely to be concentrated in areas with low access to PEP. Even given only 31 clinics, we estimate that current provisioning averts `r glue("{trans_perc(min(prop_averted))} - {trans_perc(max(prop_averted))}%")` of deaths that would occur due to rabies in the absence of PEP. We estimate that expanding PEP access has the potential to reduce deaths due to rabies, with expansion to an additional 250 clinics potentially reducing burden by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 250]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == 250]))}%")`.

However, our model predicts that geographic expansion of PEP access alone is unlikely to eliminate deaths due to rabies, and that even given expansion of PEP to all major health centers (n = 1676) in the country, between `r glue("{round(min(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))} - {round(max(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))}")` deaths may still occur nationally.

This result is driven by three key factors. First, given the baseline levels of reported bite incidence and our assumptions of underlying rabies exposure, we estimate that burden in areas with low travel times could still potentially be high {inline estimate of reporting/incidence}. Second, given the landscape of clinics and travel times across the country, even with expansion to all major health centers in the country, travel times remain high for a significant proportion of the population {inline estimate of prop pop living > 3 hrs}. Finally, the travel time effect, while consistently non-zero and negative, does not account much of the variance observed in the data for locations of similar travel times, suggesting that other factors may be driving reported bite incidence not included in our study {inline overdispersion of sigma_e}.

We find that after a certain number of clinics are added, rather than large increases in reported incidence, we see shifts in the clinic locations where bites are reported to. Vial demand reflects these patterns as well, however, vials needed per death averted match predictions from a recent cost effectiveness analysis of PEP expansion, suggesting that expanding PEP access will remain highly cost effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019].

>>>>

```{r setupmain, include=FALSE}
# Libraries
library(knitr)
library(tidyverse)
library(data.table)
library(rgdal)
library(glue)
library(here)
library(magrittr)

@@

Using the approach described in @weissGlobalMapTravel2018, we used GPS points of ARMC to estimate the travel time to the nearest clinic provisioning PEP at an approximately 1 x 1 km scale. We then extracted the mean travel times as well as the mean of travel times weighted by population for the district and commune (the administrative unit below the district) level. We also compared these metrics to euclidean distance, taking the minimum distance from the administrative unit centroid to any ARMC. We used administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs.

@@

We validated this approach to estimating form undersubmission by comparing estimated vial demand given the total reported bites (corrected for under-submission) for the years 2014-2017 to actual vials provisioned to clinics over this same period. Vial demand was simulated under simplified assumptions of PEP administration and adherence[@rajeevHealthcareUtilizationProvisioning2018], assuming that patients reported randomly across the year and then completed doses 3, 7, and 28 days subsequent from the date reported. For this analysis, we look at the range of vial estimates assuming that all patients complete 3 vs. 4 doses total. During this period, the Thai Red Cross Intradermal regimen was used across Madagascar, with 0.2 mL administered per patient at each visit. Vials can be shared within a day between two patients, resulting in 0.1 background wastage per vial shared, plus any additional wasteage from unused doses discarded at the end of the day. We did not have the date or month when vials were provisioned, so we compared estimated demand to the total vials provisioned across this four year period.

@@

## Modeling reported bites as a function of travel time to closest ARMC {#methods-bitemods}
We modeled the number of reported bites as a function of travel time ($T$) in a poisson regression framework as follows: 
$$ \mu_i = e^{(\beta_{t}T_i \ + \ \beta_0 \ )}P_i$$
$$ y_i = Poisson(\mu_i) $$
where $y_i$ is our data on average number of bites reported to a clinic annually and $\mu_i$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC ($T_i$) and the human population size ($P_i$) (an offset which scales the incidence to the expected number of bites) for a given source location (district or commune). We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level. In order to more directly compare estimates between these two datasets, we also modeled the national data with a latent commune-level travel time covariate ($T_{j}$) so that:

@@

We tested whether the catchment random effect was capturing overdispersion in the data (i.e. variance greater than the mean--the expectation of variance given a poisson distribution) rather than any catchment specific effects by extending these models to incorporate an overdispersion parameter, $\sigma_e$ [@gelman2006data]:

@@

We fit all models in a Bayesian regression framework via MCMC using the R package ‘rjags’. We used a weakly informative prior for the intercept of the models centered around the mean of the bite incidence for the given data set, and for all other parameters we centered the priors around zero ($norm(0, 10^{-6})$). For the variance terms ($\sigma_0$ and $\sigma_e$), we set uniform priors ($unif(0, 10)$). We calculated the Deviance Information Criterion (DIC, a metric of model fit to data) for each candidate model as well as potential scale reduction factors (psrf, a metric of model convergence) for each covariate in the model and the multivariate psrf for the whole model [@gelman2006data].

To test how well our models predicted the data, we sampled parameter estimates from the posterior distribution for each model to generate predictions to the data. In addition, we used the models to predict the data that were not used to fit the model (i.e. we used the estimates from models fitted to the national data to predict the Moramanga data, and used estimates from models fitted to the Moramanga data to predict the national data). Finally, to check how correcting for incomplete submission of forms affected our modeling results, we fitted our final models to the raw data uncorrected for submission (i.e. assuming that data was completely reported resulting in lower estimatates of bite incidence) and with a lower cut-off for identifying periods of undersubmission (7 days vs. 15 days, resulting in higher estimates of undersubmission and accordingly bite incidence).

@@

For $E_i$, we center the distribution at the lower end of our estimate of exposure incidence from the Moramanga District (42 exposures/100,000 persons). The range for $E_i$ is estimated using the approach used in [@whorabiesmodelingconsortiumPotentialImpactImproved2019] where rabies incidence in the dog population is estimated to be approximately 1% annually in the absence of dog vaccination, and that on average a rabid dog exposes 0.39 persons. These assumptions are derived from modeling and long term contact tracing data from the Serengeti District, Tanzania [@hampsonRabiesExposuresPostExposure2008; @hampsonSurveillanceEstablishElimination2016; @whorabiesmodelingconsortiumPotentialImpactImproved2019]. We then take the range of exposure incidence as the probability of exposure multiplied by the number of expected rabies cases in dogs assuming 1% incidence and the estimate of the dog population using a range of human to dog ratios (HDRs) of 5 (upper estimate) and 25 (lower estimate). As there is little data on dog populations in Madagascar [@andriamandimbySurveillanceControlRabies2013], this range of HDRs encompasses a wide range of observed HDRs across Africa [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating], as well as recent subnational estimates from Madagascar [@leblancclaireRabiesMadagascarThreepronged2019], and also generates similar human exposure incidences as observed previously across Africa [@beyeneEstimatingBurdenRabies2018; @hampsonRabiesExposuresPostExposure2008]. See figure S4 for further details.

@@

$B_i$, the number of total reported bites, is the predicted number of bites from the models of bite incidence as a function of travel times. To get $R_i$, the number of reported rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures, drawn from a triangular distribution with range between 0.2 - 0.6. So that rabid reported bites cannot exceed the total expected number of human rabies exposures or a maximum reporting (as we know that even with minimal travel times, people may not report for PEP for other reasons), we constrain $p_{rabid}$ as follows:
$$ p_{rabid}=\begin{cases} x, & \text{if}\ \frac{E_i\rho_{max}}{B_i} > x \\ \frac{E_i\rho_{max}}{B_i}, &\text{otherwise} \end{cases}$$
where $\rho_{max}$ is the maximum reporting. We estimate $\rho_{max}$ from the Moramanga ARMC data for the commune of Moramanga Ville, the closest commune to the ARMC (average of 3.12 minutes travel time to the clinic), and where contact tracing data estimated that approximately 2% of rabies exposure go unreported [@rajeevHealthcareUtilizationProvisioning2018]. We assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies.

We subtract $R_i$ from $E_i$, the number of total human rabies exposures drawn from a triangular distribution with range of 15.6 - 76 exposures per 100,000 persons, to get the number of unreported exposures ($U_i$). Deaths due to rabies in each administrative unit ($D_i$) are then estimated by mutliplying $D_i$ times $p_{death}$.
Deaths averted are calculated similarly as the reported rabies exposures $R_i$ that would have died in the absence of PEP given $p_{death}$ (i.e $A_ip_{death}$). Table 1 lists all parameter values and sources used in the decision tree. \

@@

```{r ncsbs}
## candidate points
csbs <- fread(here("data/processed/clinics/csb2.csv"))
clin_per_comm <- fread(here("data/processed/clinics/clinic_per_comm.csv"))
clin_per_dist <- fread(here("data/processed/clinics/clinic_per_dist.csv"))
missing_comms <- fread(here("output/stats/missing_comms.csv"))
comms_no_clins <- length(mada_communes) - (nrow(clin_per_comm) + nrow(ctar_metadata) + nrow(clin_per_dist))
csbs_max <- rbind(csbs, clin_per_comm[!(clinic_id %in% csbs$clinic_id)], fill = TRUE)

We developed a framework to rank clinics by how much they could improve access for people currently underserved and to estimate the incremental reduction in burden as PEP is expanded to these clinics. We simulate expansion first to each district (N = `r length(mada_districts)`)and then to each commune in the country for all communes with a clinic. We select the CSB II (Centre de Sante Niveau II, major health centers with capacity for delivering vaccines) in the highest density grid cell of the administrative unit as candidates for expansion. At the commune level, `r nrow(missing_comms) - comms_no_clins`  communes did not have a CSB II, so we chose the CSB I (Centre de Sante Niveau I, secondary health posts) in the highest density grid cell. `r comms_no_clins` communes lacked any CSB I or II.

We first tested three different metrics for ranking clinics:

 1. The proportion of people living > 3 hours away from an existing ARMC for which travel times were reduced when a clinic is added.
 2. The proportion of people living > 3 hours away from an existing ARMC for which travel times were reduced weighted by the  magnitude of the change in travel times when a clinic is added.
 3. The mean reduction in travel times for people living > 3 hours away from an existing ARMC when a clinic is added.

For all three metrics, once any clinics added only reduced travel times below that threshold (i.e. > 3 hours away) for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` persons), we reduce the threshold by 25%.

For each of these metrics, we sequentially add the clinic with the maximal metric and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially until each district had a clinic. At each step, we estimate the incremental reduction in burden as each clinic is added using the burden framework described in the previous section. We compared the reductions in burden using the decision tree framework described previously for these three strategies, and also to 10 different scenarios where clinics were added randomly. We then use the ranking metric that reduced burden the most effectively as clinics were added to expand iteratively to all communes with a clinic. Finally, we explore a final scenario where all additional CSB II in the country provision PEP in addition to the clinics added to each district and commune for a total of `r nrow(csbs_max)`.

We also estimate how expansion to additional clinics may impact vaccine demand. Specifically, we use our model-predicted estimates of annual bites at the district and commune level and then aggregate these to the clinic level. As clinics are added, multiple clinics may serve a single district or commune. Therefore, we allocated bites to each clinic according to the proportion of the population in each administrative unit for which it was the closest. For each clinic, we simulate throughput by randomly assigning patient presentation dates, and then assume perfect compliance (i.e. all patients reporting 3 and 7 days after their first dose) to generate dates of subsequent doses. We use these dates to estimate vial usage given routine vial sharing practices in Madagascar as described previously, but under the assumption of adoption of the new abridged intradermal regimen recommended by WHO (2 injection sites on day 0, 3, and 7, [@tarantolaIntradermalRabiesPostexposure2018]). We take the mean of 1000 simulations of throughput for each catchment and set of clinics.

## Sensitivity analysis {#methods-sensitivity}
To test the effect of our model assumptions on estimates of rabies burden and vial demand, we did a univariate sensitivity analyses of both parameters from the models of bite incidence and the decision tree parameters. For each parameter in the decision tree, we simulated predcitions fixing estimates to the upper and lower range of each parameter, while keeping the other parameters unchanged (Table S6.1). For parameters from the models of reported bite incidence, we fixed the estimate at the upper and lower 95% quantiles of their posterior distributions (Table S6.2). We also look at how systematic variation in rabies incidence with human population size could affect our estimates of burden. Specifically, if we believe that human to dog ratios are positively correlated with human populations (i.e. if dog ownership/populations are higher in more populated, urban areas), then we might expect higher rabies exposure incidence within our range as population size increases. Alternatively, if we believe that the human to dog ratio is inversely correlated with human population size (i.e. if dog ownership is more common in less populated, rural areas), we might expect rabies exposure incidence within our range to scale negatively with human population size. We use scaling factors to scale incidence either positively or negatively with the population sizes observed at the district and commune levels, while still constraining them within the range of exposure incidence used in the main analyses (15.6 - 76 exposures per 100,000 persons, see Fig S6.4) and simulated baseline burden, as well as the impact of expanding PEP access.

## Data and code 
All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the [supplementary references]{#srefs}. All processed data, code, and outputs are archived at this zenodo repository (to be archived upon submission) and maintained at [this github repository](https://github.com/mrajeev08/MadaAccess). The raw bite patient data at the national level are maintained in a database by IPM. This database was last queried on September 19, 2019 for these analyses. Anonymized raw data from this database are available from IPM following insitutional data transfer protocols. Anonymized raw data collected from the Moramanga District were retrieved from the Wise Monkey Portal (http://www.wisemonkeyfoundation.org/) on the same date and are shared in the archived repository.

@@

Vial demand also increases (Fig 7B), and more vials are needed per averted death as PEP access is expanded (Fig 7C). Our model predicts an increase from `r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 0]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 0]), 2)` vials under current clinic provisioning and with the abridged Thai Red cross regimen (i.e. visits on days 0, 3, 7), to ~`r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 250]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == 250]), 2)` vials with 250 clinics, and approximately `r round(min(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` - `r round(max(1 - natl_preds_all$vials_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]), 2)` vials if all CSB IIs were to provision PEP. Importantly, clinic throughput decreases, as well, with clinics seeing fewer patients per day (S5.7), as reported bite incidence saturates (Fig S5.5), and instead there are shifts in which clinics bites are reported to.

@@

## Main findings
Travel time to the closest ARMC was a strong predictor of reported bite incidence in Madagascar. Using an adapted decision tree framework, we find that the burden of rabies is likely to be concentrated in areas with low access to PEP. Even given only 31 clinics, we estimate that current provisioning averts `r glue("{trans_perc(min(prop_averted))} - {trans_perc(max(prop_averted))}%")` of deaths that would occur due to rabies in the absence of PEP. We estimate that expanding PEP access has the potential to reduce deaths due to rabies, with expansion to one clinic provisioning PEP per district reducing deaths by `r glue("{trans_perc(min(1 - props$deaths_mean[props$scenario == 83]))} - {trans_perc(max(1 - props$deaths_mean[props$scenario == 83]))}%")`.

However, our model predicts that geographic expansion of PEP access alone is unlikely to eliminate deaths due to rabies, and that even given expansion of PEP to all major health centers (n = 1676) in the country, between `r glue("{round(min(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))} - {round(max(natl_preds_all$deaths_mean[natl_preds_all$scenario == max(natl_preds_all$scenario)]))}")` deaths may still occur nationally. This result is driven by two key factors. Given the landscape of clinics and travel times across the country, even with expansion to all major health centers in the country, travel times remain high for a significant proportion of the population {inline estimate of prop pop living > 3 hrs}. Finally, the travel time effect, while consistently non-zero and negative, does not account for much of the variance observed in the data for locations of similar travel times, suggesting that other factors may be driving reported bite incidence not included in our study {expected range of incidence at low travel times}.

Because of these factors, we find that after a certain number of clinics are added, rather than large increases in reported incidence, we see shifts in the clinic locations where bites are reported to. Vial demand reflects these patterns as well, however, vials needed per death averted match predictions from a recent cost effectiveness analysis of PEP expansion, suggesting that expanding PEP access will remain highly cost effective [@whorabiesmodelingconsortiumPotentialImpactImproved2019]. While the qualitative patterns hold, the actualy # of deaths is highly sensitive to our assumptions of rabies exposure incidence.
